No More, No Less Than Sum of its Parts: Monoids and The Algebra of Graphics, Statistics, and Interaction

### 2024-10-23 paper.Rmd
Rewrote some sections

---
title: "No More, No Less Than Sum of its Parts: The Algebra of Graphics, Statistics, and Interaction"
blinded: 0
authors: 
- name: Adam Bartonicek
  affiliation: Department of Statistics, The University of Auckland
- name: Simon Urbanek
  affiliation: Department of Statistics, The University of Auckland
- name: Paul Murrell
  affiliation: Department of Statistics, The University of Auckland
keywords:
- Interactive data visualization, category theory, abstract algebra, the Gestalt Principles, descriptive statistics
abstract: |
  Interactive data visualization has become a staple of modern data presentation. Yet, despite its growing popularity, there still remains a lack of a formal framework for turning raw data into summary statistics that can then be displayed by interactive graphics. This gap may stem from a subtle yet profound issue: while we would often like treat statistics, graphics, and interaction as independent entities, they are in fact deeply connected. This paper aims to shed light on this interdependence by relating it to two fundamental concepts from category theory: groups and monoids. Specifically, if we want our graphics to support interactive features which break the data into multiple parts and then combine them back together (such as in the case of linked selection), then we need the statistics in our plots to fulfill certain algebraic properties. Ultimately, understanding this hidden structure may empower data visualization experts to build more flexible and expressive interactive data visualization systems.   
bibliography: references.bib
biblio-style: apalike
nocite: |
  @magritte1929
  @horst2020
output:
  rticles::asa_article:
    includes:
      in_header: "preamble.tex"
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{tikz-cd}
---

@@

The essence of data visualization is comparison. When we visualize, we split our data into parts, summarize each part one or more summary statistics, encode the summaries into visual attributes such as position, size, or color, and finally use these attributes to draw geometric objects [@bertin1983; @wilkinson2012; for a recent review, see @franconeri2021; @wilke2019]. For example, when drawing a typical barplot, we split our data into parts based on the levels of some categorical variable, count the number of cases in each part, and finally represent the counts by drawing bars of the corresponding heights. In this way, each bar becomes a visual representation of its own "small data set", a subset of the original data.

To ground this idea in a more concrete example that will be used throughout this paper, imagine going through your bookshelf, sorting the books by genre, and stacking them on top of each other. You would construct a physical version of the barplot, forming one "bar" (stack) per genre: one bar for sci-fi novels, one bar for romance, one bar for detective fiction, and so on. Even from this very simple example, interesting mathematical properties jump out: for example, a single book cannot go into two stacks (the subsets of the data are *disjoint*) and if all of the stacks are combined back together, we should end up with the bookshelf we started with [the subsets are *surjective* and cover the entirety of the data set, see @ziemkiewicz2009].

Now, just as we can learn about the contents of our library by organizing books in the way we just described, we can learn about our data by splitting it into parts. A natural question then arises: can we learn more by subdividing our data into even smaller parts and comparing these? For instance, in our book example, we can split each stack of books into two smaller stacks based on the gender of the author. Each smaller stack will then represent the unique combination (Cartesian product) of the two categorical variables (genre and gender). Moreover, there arises a hierarchical relationship between the stacks: by combining two genre-gender stacks (child components) together, we can recover the original genre stack (parent component). Similar situation often arises in interactive graphics, where we use techniques such as linked selection to highlight specific subsets of our data. Either way, by forming these finer partitions, we can represent our data with greater granularity and discover trends that may otherwise hide in the aggregate.

However, once we have subdivided each category into its child components, we need to represent the information in a way that visually preserves the hierarchical nature of this relationship. In data visualization, there are two popular methods for doing this, known as "stacking" and "dodging" (see Figure \ref{fig:playfair}). These methods have existed at least since the time of William Playfair [-@playfair1801; -@playfair1822]. Let's briefly illustrate them on the example of the classical barplot. In a stacked barplot, bar segments representing the product of the two factor variables are plotted vertically on top of each other ("stacked"; the operation can be also thought of as "highlighting" parts of the bar). In contrast, in a dodged barplot, bars are plotted side-by-side, as clusters of bars. Finally, there is also a third popular method, known as layering, where we plot bar segments in separate graphical layers and use partial transparency to mitigate overplotting (this technique may have been a bit challenging for William Playfair but has been made simple with computers).

```{r playfair}
#| echo: false
#| fig-align: "center"
#| fig-cap: "Examples of the three methods for dealing with bars that have subdivided into segments. Top row: dodging (left) and stacking/highlighting (right); both by William Playfair (1801, 1822). Bottom row: layering (created with ggplot2; Wickham, 2010)."
#| message: false
#| warning: false

@@

p3 <- ggplot(mpg, aes(manufacturer, fill = factor(year))) +
  geom_bar(position = "identity", alpha = 0.5) +
  scale_x_discrete(labels = abbreviate) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()
)

@@

Much has been written about the relative merits of stacking, dodging, and layering. For example, layering is only useful with few categories, as blending many colors can make it difficult to tell the categories apart [@franconeri2021; @wilke2019]. Further, in a landmark study, @cleveland1984 showed that people tend to be less accurate when reading information from stacked bar charts as opposed to dodged bar charts. Specifically, since the lower y-axis coordinate of a stacked segment is pushed up by the cumulative height of the segments below it, it becomes difficult to accurately compare segments' length, both within and across bars [@cleveland1984]. Subsequent research has independently validated these findings and expanded upon them [see e.g. @heer2010; @thudt2016; @quadri2021]. Ultimately, the suboptimal statistical legibility of stacking has led some authors to advise against its use altogether [@kosara2016; @wilke2019], or at least urge caution [see e.g. @byron2008; @cairo2014; @franconeri2021].

@@

We argue that not only is stacking a useful technique, but that it can in fact offer valuable insights about the mathematical structure underlying our visualizations. To make the strongest case possible, we bring forward evidence from several levels of the visualization process. We start by discussing the properties of visual perception and how they apply to both static and interactive visualizations. Next, we explore the process of computing statistical summaries that form the basis of our plots. Finally, we dive into the algebraic properties of the functions underlying these statistical summaries. Throughout the text, we use the example of the humble barplot to ground our discussion.

Our goal is to convince you that, in order to produce coherent and efficient interactive graphics, we simply cannot treat graphics, statistics, and interaction as independent components. Instead we need to consider them holistically, within the context of their respective visual and mathematical properties. Understanding this deep web of interdependence may in turn empower us to innovate and build new kinds of robust and expressive interactive visualizations.

@@

Over the course of our evolutionary history, the human brain has evolved a sophisticated pipeline for processing visual information. This pipeline has been the subject of a great body of research in cognitive neuroscience [@goebel2004; @knudsen2020; for a brief overview, see @ware2019, pp. 20]. Many of the findings from this research are directly applicable to data visualization. For example, certain particularly salient visual stimuli are perceived automatically, rapidly (<10 ms), in parallel, and at a constant speed regardless of the number of distractors [@treisman1985]. This phenomenon, known as pre-attentive processing, is key for data visualization, as the right use of visual attributes such as color, contrast, or shape can help us design figures which communicate insights much more effectively [@ware2019].

However, many of the findings from visual perception research are quite specific; to guide effective visualization, we also need some general, high-level rules. One such set of high-level rules are the Gestalt principles of visual perception [see e.g. @cairo2012; @ware2019; @rosli2015; @vanderplas2020; @todorovic2008; @pinker1990]. These principles formulate a set of fundamental "laws" related to the perception of patterns, primarily visual. They were discovered by the founders of the Gestalt school of psychology early in the 20th century, primarily by means of self-observation, and, despite not being grounded in sound understanding of the neural mechanisms of the brain, have nevertheless been later independently supported by findings from cognitive neuroscience [@guberman2017; @ware2019; @wertheimer1938].

Among the Gestalt principles, there are a few which are particularly relevant to data visualization. First, there are the two principles of proximity and similarity. These principles state that objects that which appear in close together and/or similar in appearance tend to be perceived as part of single group. Second, the closure and common region principles state that regions with closed contours tend to be perceived as single, unified objects. Finally, the figure and ground principle states we tend to see one part of a scene as the central object (the "figure") while the rest is perceived as the background (the "ground"). This can be modulated by other Gestalt principle: for example, smaller, darker, and symmetric objects with closed contours are more likely to be perceived as the figure [@ware2019].

Whether we intend them to or not, the Gestalt principles affect how our data visualizations are perceived. As such, it is a good idea to try to use them to our advantage. Sometimes, little effort is needed. Such is the case, for example, when identifying clusters in a scatterplot [@rosli2015]: the principles of proximity and similarity ensure that clouds of similarly shaped and colored points will be perceived as discrete groups by the observer. In other cases, however, our design choices can have a significant impact. For instance, another way to have objects to be perceived as part of a group is to draw a closed shape around them [@cairo2012]. Alternatively, if we want to highlight a particular part of a data set, we can leverage the figure and ground principle and use attributes such as color and saturation to make it stand out [for specific examples, see @franconeri2021; @wilke2019, p. 34].

Finally, ignoring or violating the Gestalt principles can make our graphics less legible or even misleading. @tufte2001 had documented many such violations, such as deceptive 3D effects, inadvertent optical illusions caused by dense stripe patterns, and the overuse of superfluous graphical elements he termed "chartjunk". Similarly, @cairo2019 highlights many cases of graphics that violate the Gestalt principles. Finally, there is also the issue of well-designed graphics that communicate no statistical information at all, or even information that is blatantly false [@cairo2014; @franconeri2021; @cairo2019; @gelman2013; @tufte2001]. Whether intentional or not, poor design can have a significant impact on the legibility and effectiveness of statistical graphics.

@@

The Gestalt principles apply to both static and interactive visualizations, however, the way they apply is different. Static data visualizations can be designed well or poorly, but this design quality does not change over time. In contrast, interactive visualizations change over time, with user interaction. As such, an interactive visualization may conform to the Gestalt principles at one instance of time and violate them in another. In fact, interactive visualizations can even differ in how well they leverage design principles throughout time: the animations that interactions induce may themselves conform to good or poor design [see @hullman2013; @ware2019].

Thus, interactivity inherently places further constraints the design of figures. If we want well-designed interactive figures, we need to ensure that they conform to good design principles across all states induced by user-interaction. This may not always be easy. We will investigate this idea in the following section, on the example of linked selection.

@@

While these constraints may not seem too restrictive, we argue that they pose significant challenges for certain types of plots and visualization styles. Specifically, some kinds of plots may produce undesirable visual behavior when combined with linked selection. In fact, this may explain why stacking is preferred over dodging within the interactive data visualization community. Let's explore this idea in more depth by comparing linked selection with a stacked vs. dodged barplot.

```{r display}
#| echo: false
#| fig-height: 3.5
#| fig-cap: "\\label{fig:display}The visual coherence of stacking vs. dodging
#| during linked selection on the `mtcars` data set [@henderson1981]. Bars show
#| the numbers of cars in the 4-, 6-, and 8-cylinder categories. 
#| Dark segments represent selected cases, light segments represent unselected cases, 
#| and the plots' contour is shown via a light outline. Plots left to right show more cars being selected. 
#| Top row: the contour of a stacked barplot remains constant throughout selection
#| (three bars of the same height). Bottom row: the contour of a dodged barplot 
#| changes dramatically with selection. Note the changes to the upper y-axis limit 
#| and the absence of the lighter segment among the 6-cylinder cars
#| in the rightmost plot (since all 6-cylinder cars are selected, 
#| the darker segment has been stretched to fill the available space; 
#| `ggplot2` default for handling empty subcategories)."

@@

cols <- c('grey80', '#56B4E9', '#009E73')
mtcars$cyl <- factor(mtcars$cyl)
plots <- list()

@@

First, we propose that stacking is the superior method for *displaying* selection (see Figure \ref{fig:display}). In a stacked barplot, each stacked bar presents itself as a single visual entity, thanks to the four Gestalt principles: proximity, similarity, closure, and common region. Since the stacked segments are placed right on top of each other, and share the same width and closed border, they are perceived as part of a unified whole (the bar). This remains true throughout selection - no matter how the heights of the highlighted segments within a stacked bar change, the height of the whole bar remains constant, and so does the overall outline of the plot.

Dodged barplots lack this degree of visual consistency. In a dodged barplot, bar segments placed side-by-side in clusters, and, since each segment can be of different length, these clusters lack a shared border (common region). This leads to several design challenges. First, when selection occurs, the jagged contour of the dodged barplot can change dramatically as more cases are added or removed. This creates distracting visual clutter that makes it difficult to track changes over time. Second, since the heights of the bar segments are not bound by any value present in the plot (unlike the heights of the segments in the stacked barplot, which are bound by the height of the tallest bar), we need to decide how the handle the upper y-axis limit in response to selection. One option is to make the upper y-axis limit dynamic. If we do this, we ensure that the bar segments always fit the plotting region, however, we also lose the context that the upper y-axis limit provides each time selection occurs. Alternatively, we can assign a static value to the upper y-axis limit, however, then we risk the bar segments growing outside of the bounds of the plotting region.

@@

Stacked barplots offer several other advantages over dodged bars within the context of linked selection. First, we propose that stacking also provides a simpler mental model for *selection*. Since each stacked bar is a single visual entity, it makes intuitive sense that clicking it will select the entire category corresponding to it. Indeed, this is how many interactive data visualization systems implement linked selection: clicking on part of an object selects the category corresponding to the entire object [see e.g. @cook2023; @vegalite2022; @vanderplas2018; @xie2014; @theus2002; @urbanek2003; @urbanek2011; @swayne2003]. In dodged barplots, on the other hand, the expectation is less clear: when we click on a bar segment in a dodged barplot, will we select the entire category or just the cases corresponding to the segment? Second, in a stacked barplot, the bar width and inter-bar gap are constant throughout selection. In a dodged barplot, it is not clear how to handle the gaps when some selection groups are empty. For example, when no cases are selected, should we leave gaps for the (currently absent) highlighted segments, or draw wide bars and only shrink these in response to selection? Finally, by presenting fewer visual entities on the screen, stacked bars may also reduce cognitive load [@sweller2019; for a non-technical overview, see @knaflic2015].

To summarize, stacked bars respond gracefully to selection. In contrast, dodging produces several undesirable behaviors when combined with selection: it produces jagged objects with non-uniform outline that changes dramatically with selection. These difference may explain why stacking is still popular in the interactive data visualization community.

@@

While data visualizations are made up of geometric objects such as points, lines, or areas, these objects alone are not enough to constitute data visualizations. A coloured rectangle holds no meaning in and of itself; a canvas filled with such rectangles is just an image (see Figure \ref{fig:magritte}). To turn an image of rectangles into a plot, we need to make sure that each rectangle accurately represents some underlying data.

```{r magritte}
#| echo: false
#| fig-height: 2.5
#| fig-show: "hold"
#| fig-align: "center"
#| fig-cap: "An image is not the thing it represents. Left: The Treachery of Images by René Magritte (1929) makes the point that an image of a thing is different from the thing itself. Right: similarly, an image containing a few coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data."

@@

However, we can rarely just "throw" data at the computer screen, in its raw form. Instead, we often first need to translate it into a format that the graphics device can understand. In most data visualization packages, this is done in two steps: aggregation and scaling.

First, the data needs to be aggregated or "wrangled" into sets of statistical summaries. These summaries may be very simple: for instance, in a typical scatterplot, we assign each point the raw values of the x- and y-variables (the summary function is then just identity, $f(x) = x$). More often, however, the data is grouped and aggregated in some way. For example, as was discussed previously, in a barplot, data points are grouped by the levels of the x-axis variable, and each group is then summarized by some descriptive statistic, such as count, sum, mean, or maximum. Similarly, in a typical histogram, the x-axis variable is divided into bins, and each bin is then summarized by the number of cases that fall inside it.

Second, the computed statistics have to be encoded into graphical attributes such as position, length, size, or colour [see e.g. @franconeri2021; @wilkinson2012; @pinker1990]. This is the job of specialized functions known as scales or coordinate systems [see e.g. @murrell2005; @murrell2007; @wickham2010; @wickham2016; @wilkinson2012]. For instance, in a typical scatterplot, the values of the x-axis variable are mapped to the x-position, such that points corresponding to small values of the x-axis variable are placed near the left-hand side of the plot (and larger values are place near the right-hand side). These mappings may be subject to non-linear transformations, such as log or square-root, and may also apply to discrete values such as barplot categories. Scales and coordinate systems are a rich subject and we have barely scratched the surface here; interested reader is advised to see @wilkinson2012, @ziemkiewicz2009, or @hotz2020.

@@

What does stacking mean in the context of the statistics we are trying to represent? To a novice, stacking may appear as a purely graphical operation. If they attempt to implement a data visualization pipeline from scratch, they may leave stacking to the very end, after the statistical summaries have already been translated into screen coordinates.

This view of stacking as a purely graphical operation has a certain appeal. Many modern data visualization systems follow the grammar-based model of visualization [@mcnutt2022; @kim2022; @vanderplas2020; @wickham2010; @satyanarayan2014; @satyanarayan2016; @wilkinson2012]. This model emphasizes compositionality, allowing plots to be built from independent, modular components, which the users can combine freely. For example, in the popular `ggplot2` library [@wickham2010], plots are built by combining components such as geometric objects (called `geoms`) and statistical summaries (`stats`). These grammar-based approach offers a great deal of expressiveness, and has gained widespread popularity as a result.

Thus, it may be tempting to treat stacking just as another modular component of a visualization, such that we may be able to stack any geometric objects and statistics we like. Unfortunately, this approach is deeply flawed. Many data visualization experts have issued explicit warnings about this:

@@

Then, we find there are other mathematical operators that also behave this way. For example, the product of products or the maximum of maximums are also valid overall statistics, see Figure \ref{fig:sumprodmax}. In fact, there is an entire class of mathematical objects with this exact behavior: monoids. We will explore monoids in the following section.

@@

Even very "non-number-like" operations such as concatenation of strings can be a monoid:

@@

Another way to show the same idea is the following commutative diagram:

>>>>

---
title: "No More, No Less Than Sum of its Parts: Monoids and The Algebra of Graphics, Statistics, and Interaction"
blinded: 0
authors: 
- name: Adam Bartonicek
  affiliation: Department of Statistics, The University of Auckland
- name: Simon Urbanek
  affiliation: Department of Statistics, The University of Auckland
- name: Paul Murrell
  affiliation: Department of Statistics, The University of Auckland
keywords:
- Interactive data visualization, category theory, abstract algebra, the Gestalt Principles, descriptive statistics
abstract: |
  Interactive data visualization has become a staple of modern data presentation. Yet, despite its growing popularity, there still remains a lack of a formal framework for turning raw data into summary statistics that can then be displayed by interactive graphics. This gap may stem from a subtle yet profound issue: while we would often like treat statistics, graphics, and interaction as independent entities, they are in fact deeply connected. This paper aims to shed light on this interdependence by relating it to two fundamental concepts from category theory: groups and monoids. Specifically, if we want our graphics to support interactive features which split our data into parts and then combine these parts back together (such as linked selection), then we need the statistics in our plots to fulfill certain algebraic properties. Ultimately, understanding this hidden structure may empower data visualization experts to build more flexible and expressive interactive data visualization systems.   
bibliography: references.bib
biblio-style: apalike
nocite: |
  @magritte1929
  @horst2020
output:
  rticles::asa_article:
    includes:
      in_header: "preamble.tex"
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{tikz-cd}
---

@@

The essence of data visualization is comparison. When we visualize, we split our data into parts, summarize each part by one or more summary statistics, encode the summaries into visual attributes such as position, size, or color, and, finally, use these attributes to draw geometric objects [@bertin1983; @wilkinson2012; for a recent review, see @franconeri2021; @wilke2019]. For example, when drawing a typical barplot, we split our data into parts based on the levels of some categorical variable, count the number of cases in each part, and finally represent these counts by drawing bars of corresponding heights. In this way, each geometric object (bar) becomes a representation of its own small data set, a subset of the original data.

To ground this idea in a concrete example that will be used throughout this paper, imagine going through your bookshelf, sorting the books by genre, and stacking them on top of each other. You will construct a physical version of the barplot, forming one "bar" (stack) per genre: one bar for sci-fi novels, one bar for romance, one bar for detective fiction, and so on. Even this very simple example reveals some interesting mathematical properties. For example, a single book cannot go into two stacks (the subsets of the data are *disjoint*) and if all of the stacks are combined back together, we should end up with the bookshelf we started with [the subsets are *surjective* and cover the entirety of the data set, see @ziemkiewicz2009].

Now, just as we can learn about the contents of our library by sorting books into stacks, we can learn about our data by splitting it into parts. A natural question then arises: can we learn more by subdividing our data into even smaller parts? For instance, in our book example, we can split each stack of books into smaller stacks based on the gender of the author. Each smaller stack then represents the unique combination (Cartesian product) of the two categorical variables (genre and gender). Moreover, there arises a hierarchical relationship between the stacks: by combining genre-gender stacks (child components), we can recover the original genre stack (parent component). Similar situation often arises in interactive graphics, where we use techniques such as linked selection to highlight specific subsets of our data. Either way, by forming these finer partitions of our data, we can represent it with greater granularity and discover trends that may otherwise hide in the aggregate.

However, once we have subdivided each category into its child components, we need to represent the information in a way that visually preserves this hierarchical relationship. In data visualization, there are two popular methods for doing this, known as "stacking" and "dodging" (see Figure \ref{fig:playfair}). These methods have existed at least since the time of William Playfair [-@playfair1801; -@playfair1822]. Let's briefly illustrate them on the example of the barplot. In a stacked barplot, bar segments representing the product of the two factor variables are plotted vertically on top of each other ("stacked"; the operation can be also thought of as "highlighting" parts of the bar). In contrast, in a dodged barplot, bars are plotted side-by-side, as "clusters". Finally, there is also a third more recent method, known as layering, where we plot bar segments in separate graphical layers and use partial transparency to mitigate overplotting.

```{r playfair}
#| echo: false
#| fig-align: "center"
#| fig-cap: "Examples of three methods for dealing with bars that have split into child components (segments). Top row: dodging (left) and stacking/highlighting (right); both by William Playfair (1801, 1822). Bottom row: layering (created with ggplot2; Wickham, 2010)."
#| message: false
#| warning: false

@@

p3 <- ggplot(mpg, aes(manufacturer, fill = factor(year))) +
  geom_bar(position = "identity", alpha = 0.5) +
  scale_x_discrete(labels = NULL) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()
)

@@

Much has been written about the relative merits of stacking, dodging, and layering. For example, layering is only useful with few categories, as blending many colors can make it difficult to tell the categories apart [@franconeri2021; @wilke2019]. Further, in a landmark study, @cleveland1984 showed that people tend to be less accurate when reading information from stacked bar charts as opposed to dodged bar charts. Specifically, since the lower y-axis coordinate of a stacked segment is pushed up by the cumulative height of the segments below, it becomes difficult to accurately compare segments' length, both within and across bars [@cleveland1984]. Subsequent research has independently validated these findings and expanded upon them [see e.g. @heer2010; @thudt2016; @quadri2021]. Due to this suboptimal statistical legibility, many data visualization researchers have urged caution about stacking [see e.g. @byron2008; @cairo2014; @franconeri2021], and some have even discouraged its use altogether [@kosara2016; @wilke2019].

@@

We argue that not only is stacking a useful technique, despite its limitations, but that it can also offer valuable insights into the mathematical structure underlying our visualizations. To make the strongest case possible, we bring forward evidence from several levels of the visualization process. We start by discussing the properties of visual perception and how they apply to both static and interactive visualizations. Next, we explore the process of computing statistical summaries that form the basis of our plots. Finally, we dive into the algebraic properties of the functions underlying these statistical summaries. Throughout the text, we use the example of the humble barplot to ground our discussion.

Our goal is to convince you that, in order to produce coherent and well-behaved interactive visualizations, we simply cannot treat graphics, statistics, and interaction as independent components. Instead we need to consider them holistically, within the context of their respective visual and mathematical properties. Understanding this deep web of interdependence may in turn empower us to innovate and build new kinds of robust and expressive interactive visualizations.

@@

Over the course of our evolutionary history, the human brain has evolved a sophisticated network for processing visual input. This visual processing pipeline has been the subject of a great body of research in cognitive neuroscience [@goebel2004; @knudsen2020; for a brief overview, see @ware2019, pp. 20], and many of its findings are directly applicable to data visualization. For example, certain particularly salient visual stimuli are perceived automatically, rapidly (<10 ms), in parallel, and at a constant speed regardless of the number of distractors [@treisman1985]. This phenomenon, known as pre-attentive processing, is key for data visualization, as the right use of visual attributes such as color, contrast, or shape can help us design figures which communicate insights much more effectively [@ware2019].

However, many of the visual perception findings are quite specific and niche; to guide effective visualization, we also need some general, high-level rules. One such set of high-level rules are the Gestalt principles of visual perception [see e.g. @cairo2012; @ware2019; @rosli2015; @vanderplas2020; @todorovic2008; @pinker1990]. These principles formulate a set of fundamental "laws" related to the perception of (primarily visual) patterns. They were discovered by the founders of the Gestalt school of psychology early in the 20th century, primarily by means of self-observation, and, despite not being grounded in sound understanding of the neural mechanisms of the brain, have nevertheless been later independently supported by findings from cognitive neuroscience [@guberman2017; @ware2019; @wertheimer1938].

Among the Gestalt principles, there are a few which are particularly relevant to data visualization. First, there are the two principles of proximity and similarity. These principles state that objects that which appear in close together and/or similar in appearance tend to be perceived as part of single group. Second, the closure and common region principles state that regions with closed contours tend to be perceived as single, unified objects. Finally, the figure and ground principle states that we tend to perceive one part of a visual scene as the central object (the "figure") and the rest as background (the "ground"). This can be modulated by other Gestalt principle: for example, smaller, darker, and symmetric objects with closed contours are more likely to be perceived as the figure [@ware2019].

Whether we intend them to or not, the Gestalt principles affect how our data visualizations are perceived. As such, it is a good idea to try to use them to our advantage. Sometimes, little effort is needed. Such is the case, for example, when identifying clusters in a scatterplot [@rosli2015]: the principles of proximity and similarity ensure that clouds of similarly shaped and colored points will be perceived as discrete groups. In other cases, however, our design choices can have a significant impact. For instance, another way to make objects be perceived as part of a group is to draw a closed shape around them [@cairo2012]. Alternatively, if we want to highlight a particular part of a data set, we can leverage the figure and ground principle and use attributes such as color and saturation to make it stand out [for specific examples, see @franconeri2021; @wilke2019, p. 34].

Finally, ignoring or violating the Gestalt principles can make our graphics less legible or even misleading. @tufte2001 had documented many such violations, such as deceptive 3D effects, inadvertent optical illusions caused by dense stripe patterns, and the overuse of superfluous graphical elements termed "chartjunk". Similarly, @cairo2019 highlights many cases of graphics that violate the Gestalt principles. Finally, there is also the issue of well-designed graphics that communicate no statistical information at all, or even information that is blatantly false [@cairo2014; @franconeri2021; @cairo2019; @gelman2013; @tufte2001]. Whether intentional or not, poor design can have a significant impact on the legibility and effectiveness of statistical graphics.

@@

The Gestalt principles apply to both static and interactive visualizations, however, the way they apply is different. Static data visualizations can be designed well or poorly, but this design quality does not change over time. In contrast, interactive visualizations change over time, with user interaction. As such, an interactive visualization may conform to the Gestalt principles at one instance of time and violate them in another. In fact, interactive visualizations can even differ in how well they leverage design principles *across* time: the animations that interactions induce may themselves conform to good or poor design [see @hullman2013; @ware2019].

Thus, interactivity inherently places additional constraints the design of our figures. If we want well-designed interactive figures, we need to ensure that they conform to good design principles across all states induced by user-interaction. This may not always be easy. We will investigate the design challenged posed by interaction in the following section, on the example of linked selection.

@@

While these constraints may not seem too restrictive, we argue that they pose significant challenges for certain visualization styles. Specifically, some kinds of plots may produce undesirable visual behavior when combined with linked selection, and this may in fact explain why stacking is preferred over dodging within the interactive data visualization community. Let's explore this idea in more depth by comparing linked selection with a stacked vs. dodged barplot.

```{r display}
#| echo: false
#| fig-height: 3.5
#| fig-cap: "\\label{fig:display}The visual coherence of stacking vs. dodging
#| during linked selection. Dark segments represent selected cases, 
#| light segments represent unselected cases, 
#| and the plots' contour is shown via a light outline. 
#| Plots left to right show more cases being selected. 
#| Top row: the contour of a stacked barplot remains constant throughout selection
#| (three bars of the same height). Bottom row: the contour of a dodged barplot 
#| changes dramatically with selection. Note the changes to the upper y-axis limit 
#| and the absence of the lighter segment among the bar representing category B
#| in the rightmost plot (the darker segment has been stretched to fill the available space; 
#| `ggplot2` default for handling empty subcategories). Data source: @henderson1981."

@@

cols <- c('grey80', '#56B4E9', '#009E73')
mtcars$cyl <- factor(mtcars$cyl)
levels(mtcars$cyl) <- c("A", "B", "C")
plots <- list()

@@

First, we propose that stacking is the superior method for *displaying* selection (see Figure \ref{fig:display}). Thanks to the four Gestalt principles of proximity, similarity, closure, and common region, in a stacked barplot, each bar presents itself as a single visual entity. In other words, since the stacked segments are placed right on top of each other, and share width and a closed border, they are perceived as part of a unified whole (the bar). This remains true throughout selection - no matter how the heights of the highlighted segments change, the height of the whole stacked bar remains constant, and so does the overall outline of the plot.

Dodged barplots lack this degree of visual consistency. In a dodged barplot, bar segments placed side-by-side in clusters, and, since each segment can be of different length, these clusters lack a shared border (common region). This leads to several design challenges. First, when selection occurs, the jagged contour of the dodged barplot can change dramatically as cases are added or removed. This creates distracting visual clutter that makes it difficult to track changes over time. Second, since the heights of the bar segments are not bound by any value present in the plot (unlike the heights of the segments in the stacked barplot, which are bound by the height of the tallest bar), it is not clear how to handle the upper y-axis limit. On one hand, we could make the upper y-axis limit dynamic and have it change in response to selection. If we do this, we ensure that the bar segments always fit the plotting region, however, we also lose the context that the limit provides each time selection occurs. Alternatively, we can set the limit to a static value, however, then we are at the risk of the bar segments growing outside of the bounds of the plotting region.

@@

Stacked bars may offer several other advantages over dodged bars within the context of linked selection. First, we propose that stacking also provides a simpler mental model for *selection*. Since each stacked bar is a single visual entity, it makes intuitive sense that clicking it will select the entire category corresponding to it. Indeed, this is how many interactive data visualization systems implement linked selection: clicking on part of an object selects the category corresponding to the entire object [see e.g. @cook2023; @vegalite2022; @vanderplas2018; @xie2014; @theus2002; @urbanek2003; @urbanek2011; @swayne2003]. On the other hand, with dodged bars, the expectation is less clear: when we click on a segment in a dodged barplot, do we select the entire category or just the cases corresponding to the segment? Second, in a stacked barplot, the bar width and inter-bar gap are constant throughout selection. In a dodged barplot, it is not clear how to handle the gaps when some selection groups are empty - should we leave gaps for the (currently absent) highlighted segments, or draw wide bars and only shrink them in response to selection? If we do the former, the barplot may show asymmetric gaps and illusory clustering due to the absence of some segments. If we do the latter, we break the principle that the area of the bar should be proportional to the number of cases it represents [@cleveland1985]. Finally, by presenting fewer visual entities on the screen, stacked bars may also reduce cognitive load [@sweller2019; for a non-technical overview, see @knaflic2015].

To summarize, stacked bars exhibit a more graceful and visually pleasing behavior when combined with selection. In contrast, dodging produces non-uniform objects with jagged outline that changes significantly with selection. These difference may explain why stacking is still popular in the interactive data visualization community.

@@

Data visualizations are made up of geometric objects such as points, lines, or areas. However, these objects alone hold no meaning. Without context, an image filled with coloured rectangles is just that - an image (see Figure \ref{fig:magritte}). To turn the into a plot, we need to make sure that each rectangle accurately represents some underlying data.

```{r magritte}
#| echo: false
#| fig-height: 2.5
#| fig-show: "hold"
#| fig-align: "center"
#| fig-cap: "A representation is not the thing itself. Left: The Treachery of Images by René Magritte (1929) makes the point that a painting of a pipe is not the thing it represents (a pipe). Right: similarly, an filled with coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data."

@@

Thus, we always have to start with the data. But the raw data is rarely something we can just "throw" at the computer screen. Instead, we often first need to massage and translate it into a format that can be understood by the graphics device. In most data visualization packages, this is done in two steps: aggregation and scaling.

First, the data needs to be aggregated or "wrangled" into sets of statistical summaries. These summaries may be very simple: for instance, in a typical scatterplot, we assign each point the raw values of the x- and y-variables (the summary function is then just identity, $f(x) = x$). More often, however, the data is grouped and aggregated in some way. For example, as was discussed previously, in a barplot, data points are grouped by the levels of the x-axis variable, and each group is then summarized by some descriptive statistic, such as count, sum, mean, or maximum. Likewise, in a typical histogram, the x-axis variable is divided into bins, and each bin is then summarized by the number of cases that fall inside it.

Second, the computed statistics have to be encoded into graphical attributes such as position, length, size, or colour [see e.g. @franconeri2021; @wilkinson2012; @pinker1990]. This is the job of specialized functions known as scales or coordinate systems [see e.g. @murrell2005; @murrell2007; @wickham2010; @wickham2016; @wilkinson2012]. For instance, in a typical scatterplot, the values of the x-axis variable are mapped to the x-position, such that points corresponding to small values of the x-axis variable are placed near the left-hand side of the plot (and larger values are place near the right-hand side). These mappings may be subject to non-linear transformations, such as log or square-root, and may also translate discrete values, such as barplot categories. Scales and coordinate systems are a rich subject and we have barely scratched the surface here; interested reader is advised to see @wilkinson2012, @ziemkiewicz2009, or @hotz2020.

@@

> “This system cannot produce a meaningless graphic, however. This is a strong claim, vulnerable to a single counter-example. It is a claim based on the formal rules of the system, however, not on the evaluation of specific graphics it may produce.” 
>
> “Some of the combinations of graphs and statistical methods may be degenerate or bizarre, but there is no moral reason to restrict them.”
>
> @wilkinson2012, The Grammar of Graphics, pp. 15 and 112.

But what does stacking mean in the context of the statistics we are trying to represent? To a data visualization novice, stacking may appear as a purely graphical operation. If they attempt to implement a data visualization pipeline from scratch, they may leave stacking to the very end, after the statistical summaries have already been translated into screen coordinates.

The view of stacking as a purely graphical operation has its appeal. Many modern data visualization systems follow the grammar-based model of visualization [@mcnutt2022; @kim2022; @vanderplas2020; @wickham2010; @satyanarayan2014; @satyanarayan2016; @wilkinson2012]. This model emphasizes compositionality, and enables the user to build plots from independent, modular components. For example, in the popular `ggplot2` library [@wickham2010], plots are built by combining components such as geometric objects (called `geoms`) and statistical summaries (`stats`). This grammar-based approach offers a great deal of flexibility and expressiveness, and has gained widespread popularity as a result.

Thus, it may be tempting to treat stacking just as another modular component of a visualization, such that we may be able to freely stack any geometric objects and statistics we like. Unfortunately, there is a deep flaw in this approach. Many data visualization experts have warned about this:

@@

Then, we find there are other mathematical operators that satisfy this condition. For example, the product of products or the maximum of maximums are also valid overall statistics, see Figure \ref{fig:sumprodmax}. In fact, there is an entire class of mathematical objects with this exact behavior: monoids. We will explore monoids in the following section.

@@

Even very "non-number-like" things such as strings can be a monoid:

@@

Another way to show this is the following commutative diagram:
